### 2016-09-20 paper/1_introduction.tex
tilmanns changes + shortening

In order to develop Big Data solutions, one has to solve multifaceted problems.
The necessary technology stack comprises the recording of data, the data cleaning, the meta-data generation, the representation and integration with other data sources, the analysis and modeling of the actual problem as well as the interpretation.
Each task for itself is highly complex and deserves an individual paper.
For us, the data analysis part is the key aspect in solving Big Data problems.
Henceforth, we will concentrate on how to gain valuable insights from a huge data set.

Once big data sets were amassed, people quickly recognized that these sets contain valuable information they only have to harness.
Unfortunately, the search for useful patterns and peculiarities strongly resembled the search for a needle in a haystack.
The quest requires sophisticated tools being able to analyze the vast amounts of data.
Most of these tools are based on statistics to extract interesting features.
It is beneficial to apply these statistical means to the complete data set instead of smaller chunks.
Even if the chunks cover the complete data set, important correlations between data points might get lost in smaller parts, if treated individually.
Moreover, the statistical tools improve their descriptive and predictive results by getting fed more data.
The more data is available, the more likely it is that noise cancels out and that the significant patterns manifest.
However, these benefits come at the price of an increased computing time, which requires fast computer systems to make computations feasible.

The insights gained from collected data already help to govern business decisions, improve life quality or simply create new industries from scratch.
For example, retail stores analyze their sales, customer, pricing and weather data in order to decide which products to offer or when to do a discount sale.
That not only increases the revenue of the shops but also boosts the satisfaction of the customers by getting better offers. 
Police departments try to detect probable crime sites by extracting patterns from previously recorded criminal acts and then reinforce the policemen in that region~\cite{lohr:yt2012a}.
The intelligent deployment of policemen improves security for citizens without having to hire more officers.
Hospitals analyze their patients' records and scientific studies in order to find the cancer treatment best complying with the specifics of the patient.
The individual therapy maximizes the chance of cure~\cite{watson:2013a}.
Another example showing the benefits of data analysis made it even into a Hollywood film.
The film Moneyball is based on the true story of the Oakland Athletics baseball team, whose management built an elite team in an unfavorable financial situation.
They employed sophisticated data analysis methods to spot underestimated players, who they could cheaply recruit.
That marked the start of sports statistics, which are nowadays a common tool for professional teams.
These examples emphasize the importance and utility of information gained with analytic tools from gathered data.
Interestingly, the IDC, an international market research firm, estimates that only $0.5\%$ of the globally collected data is harnessed~\cite{gantz:iaf2012a}.
They further state that about $23\%$ of today's data is worth being analyzed.
Thus, there is still potential left for improvements.

What are the reasons for the huge gap between actual and dormant exploitation?
One of the reasons is that we are lacking the tools to keep up the pace of how fast data is created and collected.
Our analyzing methods do not scale well for the vast data sets.
The more data a system has to process, the longer it will take to finish.
Since the data is growing at an exponential rate, our analytic capacities have to improve at a similar speed to keep computations feasible.

In order to decrease the runtime of our analytic tools there are two adjusting screws.
First of all, there are the algorithms.
By finding an algorithm with a lower runtime complexity than an existing algorithm, e.g., for clustering, we can drastically decrease the required time.
However, it is strongly doubted that it is possible to develop better algorithms for certain problems.
In some cases, it is even proven that there exists a lower bound for any algorithm solving the problem.
Thus, certain problems have an inherent limitation of how fast they can be computed.

The other way to speed up the analytic tools is to make the computer faster by vertical or horizontal scaling.
To scale vertically means that we add more resources to a single computer.
For example, the main memory or the frequency of a computer could be increased, giving them more computational power.
In contrast to that, horizontal scaling means to add more computer nodes to a system.
By having more than one node, the work can be split up and distributed across the nodes.
Since each split is smaller than the original problem, the computer can process it faster.

In recent years, we have seen that clock rates of CPUs stagnated.
Before, there was a simple receipt to increase the computing power of micro-controllers; increase the clock rate, which demands more power, and shrink the channel widths to mitigate for the increased power consumption.
However, the shrinkage induces the problem of leakage, which increases the power demand again.
The consumed power is limited by the amount of energy that can be dissipated and thus there is a technological limit for the increase of clock rate.
When it became clear that the micro-controller would hit this so-called ``power wall'', one duplicated the micro-controller's functionality to support simultaneous execution of multiple applications and to harness the inherent parallelism of programs.

The emerging multi-core and distributed systems pose new challenges for programmers, since now they have to know about locking, deadlocks, race-conditions and inter-process communication in order to make most of the available hardware.
Since they have to be able to reason about interwoven parallel control flows, parallel program development is highly cumbersome and error-prone.
Therefore, new programming models are conceived, a development that relieves the programmer from the tedious low-level tasks related to parallelization such as load-balancing, scheduling of parallel tasks and fault recovery.
The programmer can concentrate on the actual algorithm and the goal he wants to achieve by using these new paradigms.

These are the reasons why Google's MapReduce~\cite{dean:c2008a} framework and its open source re-implementation Hadoop~\cite{hadoop:2008a} became so popular among scientists as well as engineers.
MapReduce is a programming framework for concurrent computations on vast amounts of data running on a large cluster.
Its ingenious idea was to separate the computation into a \emph{map} and a \emph{reduce} phase.
In the map phase, the input data set is split into elements which are all processed independently.
Afterwards, the results of the mapper are grouped together and each produced group is given to a reducer.
The reducer knows about all elements in his group and produces the final result.
The strengths of MapReduce are that it is expressive enough to implement a multitude of different algorithms while facilitating at the same time parallel execution.

But still, MapReduce and other frameworks force the user to express the program in a certain way, which is often not natural or intuitive to a user coming from a different domain.
Especially, in the field of data analytics and machine learning programs are usually expressed in a mathematical form.
Therefore, systems such as MATLAB~\cite{matlab} and R~\cite{r:1993a} are widely used and recognized for their fast prototyping capabilities and their extensive mathematical libraries.
However, these linear algebra systems lack proper support for automatic parallelization on large clusters and are thus restricting the user to a single workstation.
Therefore, the amount of processible data is limited to the size of the main memory, which constitutes a serious drawback for real-world applications.

As a solution we propose Gilbert, a distributed sparse linear algebra environment.
Gilbert allows to write MATLAB-like code and execute it on massively parallel dataflow systems.
Its expressiveness allows to quickly develop scalable algorithms to analyze web-scale data.
We introduce a novel fixed-point operator which replaces loops by recursion and which can be efficienlty parallelized.
Furthermore, we demonstrate that dataflow optimizers can be used to automatically select an optimal matrix multiplication strategy.

The rest of the work is structured as follows.
In \cref{sec:overview} we give an overview of Gilbert.
Then, in \cref{sec:gilbertFeatures}, Gilbert's language and its features are described.
\cref{sec:gilbertRuntime} presents how linear algebra operations are mapped to a dataflow system.
Gilbert's performance and scalability is evaluated in \cref{sec:evaluation}.
Related work is covered in \cref{sec:relatedWork} before the work is concluded in \cref{sec:conclusion}.

>>>>

In order to develop Big Data solutions, one has to solve multifaceted problems. %The necessary technology stack comprises the recording of data, the data cleaning, the meta-data generation, the representation and integration with other data sources, the analysis and modeling of the actual problem as well as the interpretation. Each task for itself is highly complex and deserves an individual paper.
%For us, the data analysis part is the key aspect in solving Big Data problems. Henceforth, we will concentrate on how to gain valuable insights from a huge data set. Once big data sets were amassed, people quickly recognized that these sets contain valuable information they only have to harness. 
In this work, we focus on the data analysis part of the problems. Unfortunately, the search for useful patterns and peculiarities strongly resembled the search for a needle in a haystack. The quest requires sophisticated tools being able to analyze the vast amounts of data. Most of these tools are based on statistics to extract interesting features. It is beneficial to apply these statistical means to the complete data set instead of smaller chunks. Even if the chunks cover the complete data set, important correlations between data points might get lost in smaller parts, if treated individually. Moreover, the statistical tools improve their descriptive and predictive results by getting fed more data. The more data is available, the more likely it is that noise cancels out and that the significant patterns manifest. However, these benefits come at the price of an increased computing time, which requires fast computer systems to make computations feasible. 
%The insights gained from collected data already help to govern business decisions, improve life quality or simply create new industries from scratch.
%For example, retail stores analyze their sales, customer, pricing and weather data in order to decide which products to offer or when to do a discount sale. That not only increases the revenue of the shops but also boosts the satisfaction of the customers by getting better offers. 
%Police departments try to detect probable crime sites by extracting patterns from previously recorded criminal acts and then reinforce the policemen in that region~\cite{lohr:yt2012a}.
%The intelligent deployment of policemen improves security for citizens without having to hire more officers.
%Hospitals analyze their patients' records and scientific studies in order to find the cancer treatment best complying with the specifics of the patient. The individual therapy maximizes the chance of cure~\cite{watson:2013a}.
%Another example showing the benefits of data analysis made it even into a Hollywood film.
%The film Moneyball is based on the true story of the Oakland Athletics baseball team, whose management built an elite team in an unfavorable financial situation. They employed sophisticated data analysis methods to spot underestimated players, who they could cheaply recruit.
%That marked the start of sports statistics, which are nowadays a common tool for professional teams.
%These examples emphasize the importance and utility of information gained with analytic tools from gathered data.
%Interestingly, the IDC, an international market research firm, estimates that only $0.5\%$ of the globally collected data is harnessed~\cite{gantz:iaf2012a}. They further state that about $23\%$ of today's data is worth being analyzed. Thus, there is still potential left for improvements.

What are the reasons for the huge gap between actual and dormant exploitation?
One of the reasons is that we are lacking the tools to keep up the pace of how fast data is created and collected.
Our analyzing methods do not scale well for the vast data sets.
The more data a system has to process, the longer it will take to finish.
Since the data is growing at an exponential rate, our analytic capacities have to improve at a similar speed to keep computations feasible. In order to decrease the runtime of our analytic tools there are two adjusting screws. First of all, there are the algorithms.
By finding an algorithm with a lower runtime complexity than an existing algorithm, e.g., for clustering, we can drastically decrease the required time. However, it is strongly doubted that it is possible to develop better algorithms for certain problems.
In some cases, it is even proven that there exists a lower bound for any algorithm solving the problem.
Thus, certain problems have an inherent limitation of how fast they can be computed. The other way to speed up the analytic tools is to make the computer faster by vertical or horizontal scaling. To scale vertically means that we add more resources to a single computer.
For example, the main memory or the frequency of a computer could be increased, giving them more computational power.
In contrast to that, horizontal scaling means to add more computer nodes to a system.
By having more than one node, the work can be split up and distributed across the nodes.
Since each split is smaller than the original problem, the computer can process it faster. In recent years, we have seen that clock rates of CPUs stagnated.
%Before, there was a simple receipt to increase the computing power of micro-controllers; increase the clock rate, which demands more power, and shrink the channel widths to mitigate for the increased power consumption.
%However, the shrinkage induces the problem of leakage, which increases the power demand again.
%The consumed power is limited by the amount of energy that can be dissipated and thus there is a technological limit for the increase of clock rate.
%When it became clear that the micro-controller would hit this so-called ``power wall'', one duplicated the micro-controller's functionality to support simultaneous execution of multiple applications and to harness the inherent parallelism of programs.
The emerging multi-core and distributed systems pose new challenges for programmers, since now they have to know about locking, deadlocks, race-conditions and inter-process communication in order to make most of the available hardware.
Since they have to be able to reason about interwoven parallel control flows, parallel program development is highly cumbersome and error-prone.
Therefore, new programming models are conceived, a development that relieves the programmer from the tedious low-level tasks related to parallelization such as load-balancing, scheduling of parallel tasks and fault recovery.
The programmer can concentrate on the actual algorithm and the goal he wants to achieve by using these new paradigms. These are the reasons why Google's MapReduce~\cite{dean:c2008a} framework and its open source re-implementation Hadoop~\cite{hadoop:2008a} became so popular among scientists as well as engineers.
%MapReduce is a programming framework for concurrent computations on vast amounts of data running on a large cluster.
%Its ingenious idea was to separate the computation into a \emph{map} and a \emph{reduce} phase.
%In the map phase, the input data set is split into elements which are all processed independently.
%Afterwards, the results of the mapper are grouped together and each produced group is given to a reducer.
%The reducer knows about all elements in his group and produces the final result.
%The strengths of MapReduce are that it is expressive enough to implement a multitude of different algorithms while facilitating at the same time parallel execution.
But still, MapReduce and other frameworks force the user to express the program in a certain way, which is often not natural or intuitive to a user coming from a different domain.
Especially, in the field of data analytics and machine learning programs are usually expressed in a mathematical form.
Therefore, systems such as MATLAB~\cite{matlab} and R~\cite{r:1993a} are widely used and recognized for their fast prototyping capabilities and their extensive mathematical libraries.
However, these linear algebra systems lack proper support for automatic parallelization on large clusters and are thus restricting the user to a single workstation.
Therefore, the amount of processible data is limited to the size of the main memory, which constitutes a serious drawback for real-world applications.

As a solution we propose Gilbert, a distributed sparse linear algebra environment.
Gilbert allows to write MATLAB-like code and execute it on massively parallel dataflow systems.
Its expressiveness allows to quickly develop scalable algorithms to analyze web-scale data.
We introduce a novel fixed-point operator which replaces loops by recursion and which can be efficienlty parallelized.
Furthermore, we demonstrate that dataflow optimizers can be used to automatically select an optimal matrix multiplication strategy. Gilbert provides a Matlab-like language for distributed sparse linear algebra operations. Being such a system, it comprises the complete stack of functionalities necessary to implement a programming language. At first, the system has to divide the given source code into tokens. These tokens are parsed and an abstract syntax tree (AST) is created. In order to generate the intermediate representation (IR), type information is needed for all expressions. This information is extracted by the typing system. Knowing the types, the compiler can translate the Gilbert program into its IR. The IR is well suited to apply high-level transformations to the program. At last, the execution plan generator translates the IR into an execution plan which can be executed in parallel. %The layered system architecture is shown in \cref{fig:systemArchitecture}. 
The first layer is the language layer. It contains all functionality to parse the given Gilbert code and to compile it into the IR. The second layer is the intermediate layer and it receives the IR of the Gilbert code. The intermediate format is the ideal representation to apply language independent high-level transformations. The Gilbert optimizer applies several algebraic optimizations prior to the generation of the execution plan. The execution plan generator translates the optimized IR into a specific execution plan, depending on the selected execution engine. Once the program has been translated into an execution engine's specific plan, it is executed on the respective back end.

%\begin{figure}[t!]
% \centering
% \includegraphics[height=0.175\paperheight]{images/systemArchitecture.png}
% \caption{The layered system architecture of Gilbert. The language layer is responsible for parsing, typing and compiling the given Matlab code. The intermediate layer facilitates high-level optimization strategies. The runtime layer is responsible for executing the specified program in parallel.}
% \label{fig:systemArchitecture}
%\end{figure}

\textbf{Language Layer}. The language layer contains all the logic for translating Gilbert source code into the intermediate representation, see the subsequent section for more details. %Gilbert's supported language and its features is described in \cref{sec:gilbertLanguage}.
The resulting AST is attributed with type information by the typing system. The typer makes use of the Hindley-Milner algorithm to automatically infer types.%, as it is described in~\cref{sec:hmInference}. 
The type information enriched AST is then translated into a front end independent representation of linear algebra operations, namely the IR. The format Gilbert uses to represent linear algebra operations in a generalized format is described in Section~\ref{sec:intermediaterepresentation}.

\textbf{Intermediate Layer}. The intermediate layer contains the optimizer and the execution plan generator. The optimizer applies the optimizations described in Section~\ref{sec:gilbertOptimizer} to IR of the Gilbert program. Then the IR is translated into the execution engine's specific format. The linear algebra operations are translated into a dataflow graph as explained in Section~\ref{sec:LinearAlgebraOperations}. The system supports Flink and Spark as parallel execution engines.

%The rest of the work is structured as follows.
%In \cref{sec:overview} we give an overview of Gilbert.
%Then, in \cref{sec:gilbertFeatures}, Gilbert's language and its features are described.
%\cref{sec:gilbertRuntime} presents how linear algebra operations are mapped to a dataflow system.
%Gilbert's performance and scalability is evaluated in \cref{sec:evaluation}.
%Related work is covered in \cref{sec:relatedWork} before the work is concluded in \cref{sec:conclusion}.

### 2016-09-20 paper/3_gilbert_features.tex
Correct section hierarchy in gilbert_features
Shorten introduction
tilmanns changes + shortening

%!TEX root=paper.tex
\section{Gilbert's Language and Features}
\label{sec:gilbertLanguage}

Gilbert's language has been strongly inspired by MATLAB which should ease its adoption.
The language is a fully functional subset of MATLAB and its grammar is specified in \cite{masterthesis}.
The elementary data type is the matrix.
Gilbert supports arbitrary $2$-dimensional matrices whose elements can be \emph{double} or \emph{boolean}.
Vectors are not represented by a special type but instead are treated as a single column/row matrix.
Additionally, scalar \emph{double}, \emph{boolean} and string values are supported.

Gilbert also implements cell array types.
A cell array is defined using curly braces and commas to separate individual values.
The cells can be accessed by an index appended in curly braces to the cell array variable.
\Cref{lst:cellArray} shows how to define and access them.

@@

Gilbert supports the basic linear algebra operations defined on matrices and scalars.
They include among others the common operations \code{+}, \code{-}, \code{/} and \code{*}, whereas \code{*} denotes the matrix-matrix multiplication and all other operations are interpreted cellwisely.
The cellwise multiplication is indicated by a point preceding the operator.
Gilbert also supports comparisons operators such as \code{>}, \code{>=}, \code{==} and \code{\textasciitilde=}.

Besides the basic arithmetic operations, the user can also define named functions and anonymous functions.
The syntax of anonymous functions adheres to the MATLAB syntax: \code{@(x,y) x*x + y*y}.

An important aspect of MATLAB are loops.
MATLAB permits the user to express \code{for} and \code{while} loops.
However, these loops can have side effects.
The problem of parallelization of iterations with side effects is that the referenced external state has to be maintained.
This circumstance makes preprocessing and execution unnecessarily complex.
Gilbert offers a fixpoint operator \code{fixpoint}, which iteratively applies a given update function $f$ on the previous result of $f$, starting with an initial value $x$ at iteration $0$.

@@

MATLAB belongs to the class of dynamically typed languages and Gilbert requires type information for its parallel execution.
The parallel data processing systems, used to run Gilbert programs, have to know which data types are passed from one worker node to another.
Therefore, the MATLAB language has to be enriched with type information.
We infer this type information using the Hindley-Milner (HM) type system~\cite{hindley:tams1969a,milner:jcss1978a} and a slightly derived form of algorithm W~\cite{damas:1982a} for type inference.

In case that the type inference algorithm cannot properly infer the types, there has to be a way to resolve this problem.
We decided to pursue a similar approach as \cite{furr:2009a}.
\cite{furr:2009a} added type information to Ruby by adding special comments to the respective code sections without breaking it.

\subsection{Function Overloading}

MATLAB's basic operators, such as \code{+}, \code{-}, \code{/} and \code{*}, for example, are overloaded.
They can be applied to matrices, scalars as well as mixture of both data types.
That makes it very convenient to express mathematical problems, but from a programmer's point of view it causes some hardships.
Originally, HM cannot deal with overloaded functions properly, because it assumes that each function has an unique type.
In order to extend HM's capabilities, we allowed each function symbol to have a list of signatures.
In the case of \code{+}, the list of signatures would consist of 
\begin{eqnarray*}
matrix[double] \times matrix[double] &\rightarrow& matrix[double]\\
matrix[double] \times double &\rightarrow& matrix[double]\\
double \times matrix[double] &\rightarrow& matrix[double]\\
double \times double &\rightarrow& double
\end{eqnarray*}

@@

\subsection{Matrix Dimension Inference}
\label{sec:MatrixDimensionInference}

Matrices constitute the elementary data type in our linear algebra environment.
Besides its element type, a matrix is also defined by its size.
In the context of program execution, knowledge about matrix sizes can help to optimize the evaluation.
For instance, consider a threefold matrix multiplication $A\times B\times C$.
The multiplication can be evaluated in two different ways: $(A\times B)\times C$ and $A\times(B\times C)$.
For certain matrix sizes one way might be infeasible whereas the other way can be calculated efficiently due to the matrix size of the intermediate result $(A\times B)$ or $(B\times C)$.

@@

\section{Intermediate Representation}
\label{sec:intermediaterepresentation}

After parsing and typing of the source code is done, it is translated into an intermediate representation.
The additional abstraction layer allows Gilbert to apply language independent optimization in a coherent manner and independently of the actually used front end language.
This implementation aspect gives the flexibility to easily extend Gilbert to support other linear algebra languages as well.
A higher-level abstraction of the mathematical operations holds more potential for algebraic optimization, too.

The intermediate format consists of a set of operators to represent the different linear algebra operations.
Every operator has a distinct result type and a set of parameters which are required as inputs.
The set of intermediate operators can be distinguished into three categories: \emph{Creation operators}, \emph{transformation operators} and \emph{output operators}.

\subsection{Creation Operators}

Creation operators generate or load some data.
The \code{load} operator takes a path to a stored matrix on disk and the corresponding number of rows and columns as input parameters.
It then reads the stored data and loads the matrix into the program context.
The \code{eye} operator, known from MATLAB, generates an identity matrix of the requested size, number of rows and columns.
The operator also supports non quadratic result shapes.
\code{zeros} generates a matrix of the given size which is initialized with zeros.
\code{Randn} takes the number of rows and columns of the resulting matrix and the mean and the standard deviation of a Gauss distribution.
The specified Gauss distribution is used to generate random values for the resulting matrix.

\subsection{Transformation Operators}

The transformation operators constitute the main abstraction of the linear algebra operations.
They group operations with similar properties and thus allow an easier reasoning and optimization of the underlying program.

The \code{UnaryScalarTransformation} takes a single scalar value and applies an unary operation on it.
The \code{ScalarScalarTransformation} constitutes a binary operation on scalar values whereas \code{ScalarMatrixTransformation} and \code{MatrixScalarTransformation} represent a binary operation between a scalar and a matrix value.
The \code{VectorwiseMatrixTransformation} applies an operation on each row vector of the given matrix.
A vectorwise operation produces a scalar value for each row vector.
Similar to the vectorwise operations, the \code{AggregateMatrixTransformation} applies an operation to all matrix entries producing a single scalar result value.

The iteration mechanism is represented by the \code{FixpointIterationMatrix} and \\\code{FixpointIterationCellArray} operators.
The former operator is used for a fixpoint operation on matrices and the latter on cell arrays.
Both operators follow the semantics and have the same parameters as the fixpoint abstraction presented in \cref{eqn:fixpoint}.

\subsection{Writing Operators}

The writing operators are used to make the computed results accessible to the user by writing them back to disk.
There exists a writing operation for each supported type: \code{WriteMatrix}, \code{WriteScalar}, \code{WriteString}, \code{WriteCellArray} and \code{WriteFunction}.

\section{Gilbert Optimizer}
\label{sec:gilbertOptimizer}

@@

\subsection{Matrix Multiplication Reordering}

Matrix multiplications belong to the most expensive operations in linear algebra programs in terms of computational complexity as well as space complexity.
Intermediate results of successive matrix multiplications can easily exceed the memory capacity and thus rendering its computation infeasible.
However, a smart execution order can sometimes avoid the materialization of excessivly large matrices.

The best execution order of successive multiplications is the one that minimizes the maximum size of intermediate results.
In order to determine the best execution order, the optimizer first extracts all matrix multiplications with more than $2$ operands.
Then, it will calculate for each evaluation order the maximum size of all occurring intermediate results.
In order to do this calculation, the optimizer relies on the operands' automatically inferred matrix sizes, as described in \cref{sec:MatrixDimensionInference}.
At last, it will pick the execution order with the minimal maximum intermediate matrix size.

\subsection{Transpose Pushdown}

Transpose pushdown tries to move the transpose operations as close to the matrix input as possible.
Thereby, consecutive transpose operations accumulate at the inputs and unnecessary operations erase themselves.
An example is given in \cref{lst:pushdown}.

@@

By inserting $C$ into $E$, the expression $E=(A^T BD^T)^T$ is obtained. This term is equivalent to $DB^T A$.
The latter formulation contains only one transpose operation.

Usually multiple transpose operations occur because they are written for convenience reasons at various positions in the code.
Moreover, in complex programs it is possible that the programmer loses track of them or simply is unaware of the optimization potential.
Therefore, transpose pushdown might be a beneficial optimization.

>>>>

%!TEX root=paper.tex
\section{Gilbert Features}
\label{sec:gilbertFeatures}

Gilbert's language has been strongly inspired by MATLAB which should ease its adoption.
The language is a fully functional subset of MATLAB and its grammar is specified in \cite{Rohrmann2014}.
The elementary data type is the matrix. Gilbert supports arbitrary $2$-dimensional matrices whose elements can be \emph{double} or \emph{boolean}.
Vectors are not represented by a special type but instead are treated as a single column/row matrix.
Additionally, scalar \emph{double}, \emph{boolean} and string values are supported. Gilbert also implements cell array types.
A cell array is defined using curly braces and commas to separate individual values. The cells can be accessed by an index appended in curly braces to the cell array variable. \Cref{lst:cellArray} shows how to define and access them.

@@

Gilbert supports the basic linear algebra operations defined on matrices and scalars. They include among others the common operations \code{+}, \code{-}, \code{/} and \code{*}, whereas \code{*} denotes the matrix-matrix multiplication and all other operations are interpreted cellwisely. The cellwise multiplication is indicated by a point preceding the operator. Gilbert also supports comparisons operators such as \code{>}, \code{>=} \code{==} and \code{\textasciitilde=}. Besides the basic arithmetic operations, the user can also define named functions and anonymous functions. The syntax of anonymous functions adheres to the MATLAB syntax: \code{@(x,y) x*x + y*y}. An important aspect of MATLAB are loops. MATLAB permits the user to express \code{for} and \code{while} loops. However, these loops can have side effects. The problem of parallelization of iterations with side effects is that the referenced external state has to be maintained. This circumstance makes preprocessing and execution unnecessarily complex. Gilbert offers a fixpoint operator \code{fixpoint}, which iteratively applies a given update function $f$ on the previous result of $f$, starting with an initial value $x$ at iteration $0$.

@@

MATLAB belongs to the class of dynamically typed languages and Gilbert requires type information for its parallel execution.
The parallel data processing systems, used to run Gilbert programs, have to know which data types are passed from one worker node to another.
Therefore, the MATLAB language has to be enriched with type information. We infer this type information using the Hindley-Milner (HM) type system~\cite{hindley:tams1969a,milner:jcss1978a} and a slightly derived form of algorithm W~\cite{damas:1982a} for type inference. In case that the type inference algorithm cannot properly infer the types, there has to be a way to resolve this problem. We decided to pursue a similar approach as \cite{furr:2009a}. \cite{furr:2009a} added type information to Ruby by adding special comments to the respective code sections without breaking it.

%\subsubsection{Function Overloading}

\textbf{Function Overloading}. MATLAB's basic operators, such as \code{+}, \code{-}, \code{/} and \code{*}, for example, are overloaded.
They can be applied to matrices, scalars as well as mixture of both data types.
That makes it very convenient to express mathematical problems, but from a programmer's point of view it causes some hardships.
Originally, HM cannot deal with overloaded functions properly, because it assumes that each function has an unique type.
In order to extend HM's capabilities, we allowed each function symbol to have a list of signatures.
In the case of \code{+}, the list of signatures would consist of 
\begin{eqnarray*}
matrix[double] \times matrix[double] &\rightarrow& matrix[double]\\
matrix[double] \times double &\rightarrow& matrix[double]\\
double \times matrix[double] &\rightarrow& matrix[double]\\
double \times double &\rightarrow& double
\end{eqnarray*}

@@

%\subsubsection{Matrix Dimension Inference}
%\label{sec:MatrixDimensionInference}

\textbf{Matrix Dimension Inference}. Matrices constitute the elementary data type in our linear algebra environment.
Besides its element type, a matrix is also defined by its size.
In the context of program execution, knowledge about matrix sizes can help to optimize the evaluation.
For instance, consider a threefold matrix multiplication $A\times B\times C$.
The multiplication can be evaluated in two different ways: $(A\times B)\times C$ and $A\times(B\times C)$.
For certain matrix sizes one way might be infeasible whereas the other way can be calculated efficiently due to the matrix size of the intermediate result $(A\times B)$ or $(B\times C)$.

@@

\subsection{Intermediate Representation}
\label{sec:intermediaterepresentation}

After parsing and typing of the source code is done, it is translated into an intermediate representation. The additional abstraction layer allows Gilbert to apply language independent optimization in a coherent manner and independently of the actually used front end language. This implementation aspect gives the flexibility to easily extend Gilbert to support other linear algebra languages as well. A higher-level abstraction of the mathematical operations holds more potential for algebraic optimization, too. The intermediate format consists of a set of operators to represent the different linear algebra operations. Every operator has a distinct result type and a set of parameters which are required as inputs. The set of intermediate operators can be distinguished into three categories: \emph{Creation operators}, \emph{transformation operators} and \emph{output operators}.

%\subsubsection{Creation Operators}

\textbf{Creation Operators}. Creation operators generate or load some data.
The \code{load} operator takes a path to a stored matrix on disk and the corresponding number of rows and columns as input parameters.
It then reads the stored data and loads the matrix into the program context.
The \code{eye} operator, known from MATLAB, generates an identity matrix of the requested size, number of rows and columns.
The operator also supports non quadratic result shapes.
\code{zeros} generates a matrix of the given size which is initialized with zeros.
\code{Randn} takes the number of rows and columns of the resulting matrix and the mean and the standard deviation of a Gauss distribution.
The specified Gauss distribution is used to generate random values for the resulting matrix.

%\subsubsection{Transformation Operators}

\textbf{Transformation Operators}. The transformation operators constitute the main abstraction of the linear algebra operations.
They group operations with similar properties and thus allow an easier reasoning and optimization of the underlying program. The \code{UnaryScalarTransformation} takes a single scalar value and applies an unary operation on it. The \code{ScalarScalarTransformation} constitutes a binary operation on scalar values whereas \code{ScalarMatrixTransformation} and \code{MatrixScalarTransformation} represent a binary operation between a scalar and a matrix value. The \code{VectorwiseMatrixTransformation} applies an operation on each row vector of the given matrix.
A vectorwise operation produces a scalar value for each row vector. Similar to the vectorwise operations, the \code{AggregateMatrixTransformation} applies an operation to all matrix entries producing a single scalar result value. The iteration mechanism is represented by the \code{FixpointIterationMatrix} and \code{FixpointIterationCellArray} operators. The former operator is used for a fixpoint operation on matrices and the latter on cell arrays. Both operators follow the semantics and have the same parameters as the fixpoint abstraction presented in \cref{eqn:fixpoint}.

%\subsubsection{Writing Operators}

\textbf{Writing Operators}. The writing operators are used to make the computed results accessible to the user by writing them back to disk.
There exists a writing operation for each supported type: \code{WriteMatrix}, \code{WriteScalar}, \code{WriteString}, \code{WriteCellArray} and \code{WriteFunction}.

\subsection{Gilbert Optimizer}
\label{sec:gilbertOptimizer}

@@

%\subsubsection{Matrix Multiplication Reordering}

\textbf{Matrix Multiplication Reordering}. Matrix multiplications belong to the most expensive operations in linear algebra programs in terms of computational complexity as well as space complexity. Intermediate results of successive matrix multiplications can easily exceed the memory capacity and thus rendering its computation infeasible. However, a smart execution order can sometimes avoid the materialization of excessivly large matrices. The best execution order of successive multiplications is the one that minimizes the maximum size of intermediate results.
In order to determine the best execution order, the optimizer first extracts all matrix multiplications with more than $2$ operands.
Then, it will calculate for each evaluation order the maximum size of all occurring intermediate results.
In order to do this calculation, the optimizer relies on the operands' automatically inferred matrix sizes.%, as described in \cref{sec:MatrixDimensionInference}. 
At last, it will pick the execution order with the minimal maximum intermediate matrix size.

%\subsubsection{Transpose Pushdown}

\textbf{Transpose Pushdown}. Transpose pushdown tries to move the transpose operations as close to the matrix input as possible.
Thereby, consecutive transpose operations accumulate at the inputs and unnecessary operations erase themselves.
An example is given in \cref{lst:pushdown}. By inserting $C$ into $E$, the expression $E=(A^T BD^T)^T$ is obtained. This term is equivalent to $DB^T A$. The latter formulation contains only one transpose operation. Usually multiple transpose operations occur because they are written for convenience reasons at various positions in the code. Moreover, in complex programs it is possible that the programmer loses track of them or simply is unaware of the optimization potential. Therefore, transpose pushdown might be a beneficial optimization.

@@
